Highly Available Transactions: Virtues and Limitations

### 2013-03-30 writeup/intro.tex
New intro

Requirements for high availability, predictable low latency, and
simplicity of coordination protocols have led many recent distributed
database designers to renounce several properties provided by
traditional database systems~\cite{bigtable, pnuts, dynamo}. In
particular, many modern systems forgo providing ACID guarantees, which
often provide single-system image programmability but traditionally
suffer from unavailability and often expensive coordination
costs~\cite{davidson-survey}. In particular, the July 2000
announcement of the CAP Theorem~\cite{gilbert-cap}, which states that
a highly available system cannot provide recency guarantees in the
presence of network partitions, proved a rallying call for a
generation of scalable but semantically weak system designs. Over a
decade later, we observe a dearth of highly available systems
providing general-purpose multi-operation \textit{transactions} over
multiple data items with useful semantics.

While serializability, the gold standard of traditional ACID
transactions, is not achievable with high availability for general
read-write transactions~\cite{davidson-survey}, database systems have
a long tradition of providing weaker isolation and consistency
guarantees~\cite{adya, ansicritique, gray-isolation}. Due to their
concurrency and overall performance benefits, weak isolation models
are widely employed, overwhelmingly as the default setting in today's
traditional ``ACID'' and up-and-coming ``NewSQL'' databases and often
as the maximum guarantee offered (Section~\ref{sec:modernacid}). While
this weak isolation does not provide serializability for
general-purpose transactions, it is apparently strong enough to
deliver acceptable behavior to many application programmers and is
substantially stronger than the typical semantics of current highly
available systems. This raises the question: which of these semantics
can be provided while remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
is poorly understood. We have a strong understanding of weak isolation
in the single-server context from which it originated~\cite{adya,
  ansicritique, gray-isolation} as well as several techniques for
providing distributed serializability~\cite{bernstein-concurrency,
  spanner, granola, daudjee-session, calvin}. However, the current
literature does not provide guidance as to the operation of weakly
consistent transactional guarantees in a replicated
context. Accordingly, in this work, we consider the problem of
providing Highly Available Transactions (HATs), or semantic guarantees
regarding multiple operations operating on multiple data items that
are achievable with high availability and low latency, albeit with
some cost to achievable semantics. While our prior, preliminary
results indicated that some models are achievable~\cite{hat-hotos},
the literature lacks a deep understanding of what models are
available, which are not, and when this distinction matters.

This work attempts to bridge the gap between ACID properties and
highly available systems. We introduce a hierarchy of consistency
models and rigorously taxonomize their availability characteristics;
in doing so, we unify previous work on ACID properties~\cite{adya},
session guarantees~\cite{sessionguarantees}, and traditional
disributed ``register'' semantics~\cite{herlihy-art}. By identifying
the availability cost of a wide range of guarantees, this work allows
developers to weigh the costs and benefits of relying on particular
guarantees in their applications. While several models, like
serializability and Snapshot Isolation, are unachievable due to update
conflict avoidance (preventing Lost Update and Write Skew phenomena),
many other ACID guarantees are HAT compliant, such as providing Read
Committed isolation, preventing Fuzzy Read phenomena, and ensuring
atomicity of transactional updates. Several other guarantees, like
reading one's writes, fall into a new availability class---``sticky''
high availability---which provides an intermediate ground between
strict high availability and majority availability.

@@

In this paper, we make the following contributions:
\begin{myitemize}
\item We model high availability in a transactional environment,
  including traditional and sticky high availability.

\item We taxonomize ACID and distributed consistency properties
  according to their availability characteristics.

\item We analyze existing concurrency control algorithms, perform a
  case-study of an existing transactional application, and
  briefly evaluate a HAT database prototype.
\end{myitemize}

>>>>

In July 2000, the CAP Theorem~\cite{gilbert-cap} was crisply
formulated that a highly available system cannot provide strong
consistency guarantees in the presence of network partitions. This
became a rallying call for a generation of highly scalable but
semantically weak system designs. The NoSQL movement embraced this
mantra, leading to many production quality databases that eschewed
strong consistency guarantees, but could scale-out to many
servers~\cite{bigtable, pnuts, dynamo}. The consistency concept used
in the CAP theorem was proposed by the distributed systems community,
and concerned providing strong recency guarantees of a data item that
is replicated across servers.  However, the NoSQL movement implicitly
took it to imply that transactions are also unavailable in the
presence of network partitions. Over a decade later, we observe a
dearth of highly available systems providing general-purpose
multi-operation \textit{transactions} over multiple data items with
useful semantics.

 Serializability, the gold standard of traditional ACID transactions,
 is indeed not achievable with the requirement of high availability
 the in the presence of partitions~\cite{davidson-survey}. However,
 database systems have a long tradition of providing weaker isolation
 and consistency guarantees~\cite{adya, ansicritique,
   gray-isolation}. Due to their concurrency and overall performance
 benefits, weak isolation models are widely employed, overwhelmingly
 as the default setting in today's traditional ``ACID'' and
 up-and-coming ``NewSQL'' databases and often as the maximum guarantee
 offered (Section~\ref{sec:modernacid}). While this weak isolation
 does not provide serializability for general-purpose transactions, it
 is apparently strong enough to deliver acceptable behavior to many
 application programmers and is substantially stronger than the
 typical semantics of current highly available systems. This raises
 the question: which of these semantics can be provided while always
 remaining highly available?

Unfortunately, the relationship of the CAP Theorem and ACID semantics
is poorly understood. We have a strong understanding of weak isolation
in the single-server context from which it originated~\cite{adya,
  ansicritique, gray-isolation} as well as several techniques for
providing distributed serializability~\cite{bernstein-concurrency,
  spanner, granola, daudjee-session, calvin}.  However, the current
literature does not provide guidance as to the operation of weakly
consistent transactional guarantees in a replicated context. To make
matters worse, the distributed computing literature has produced a
large body of work on weak data consistency models for single
replicated objects. Accordingly, in this work, we consider the problem
of providing understanding semantic guarantees regarding multiple
operations operating on data items that are achievable with high
availability and low latency, albeit with some cost to achievable
semantics. While our prior, preliminary results indicated that some
models are achievable~\cite{hat-hotos}, the literature lacks a deep
understanding of what models are available, which are not, and when
this distinction matters.

This work attempts to bridge the gap between, one the one hand, the
plethora of isolation and consistency models proposed in the
transactions and distributed systems literature, and, on the other hand
highly available systems. We introduce a hierarchy of consistency
models and rigorously taxonomize their availability characteristics;
in doing so, we unify previous work on ACID properties~\cite{adya},
session guarantees~\cite{sessionguarantees}, and traditional
distributed ``register'' semantics~\cite{herlihy-art}.

Our main contributions are the following. We classify among the wide
array of isolation and consistency models those that are highly
available, denoting them as {\em Highly Available Transactions
  (HATs)}.  In doing so, we show that many implementations that
provide HAT-semantics are not available, but that the unavailability
is {\em not} inherent to their semantic guarantees. Thus, we provide
highly available implementations for HATs.  Furthermore, our
investigation shows, somewhat surprisingly, that besides
serializability, only Snapshot Isolation (SI) and Repeatable Reads
(RR) are unavailable. The reason for their unavailability stems from
their requirement to avoid update conflict (preventing Lost Updates
and Write Skew phenomena). Finally, we show that many of the weaker
data consistency models from distributed systems can be implemented in
a highly available manner as well. For that, the model of the system
has to be changed to include clients, which cannot be mobile, but
rather have to {\em stick} to one server, something which is implicit
in much of the previous work on distributed systems.

@@

%% In this paper, we make the following contributions:
%% \begin{myitemize}
%% \item We model high availability in a transactional environment,
%%   including traditional and sticky high availability.

%% \item We taxonomize ACID and distributed consistency properties
%%   according to their availability characteristics.

%% \item We analyze existing concurrency control algorithms, perform a
%%   case-study of an existing transactional application, and
%%   briefly evaluate a HAT database prototype.
%% \end{myitemize}
